RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2014-03-13 inst/doc/pubs/manuscript.Rmd
writing writing

@@

Mention of data sharing challenges and opportunities.  See [Drew et al 2013](10.1371/journal.pbio.1001636)

@@

The current stable version of the `RNeXML` package can be installed from the CRAN library <!-- Actually not yet -->
using the standard

@@

@@

@@

Other metadata is available

@@

- article citation information

- Summary data: taxa, number of taxa, number of trees, number of tips.

- Metadata of nodes, complete RDF extraction

This allows us to include multiple trees and character matrices to a single NeXML file.

<!-- explain OTU blocks somewhere -->

### Taxonomic identifiers

Add taxonomic identifier metadata to the OTU elements:

```{r}
nex <- add_trees(bird.orders)
nex <- nexml_taxize(nex)
```

Replicate examples from [treebase] metadata section: publication information, tree size, taxanomic coverage

`RNeXML` automatically adds basic metadata by default such as a timestamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.

@@

@@

[Temple Lang 2013, XML]: http://cran.r-project.org/web/packages/XML/
[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html
[Chamberlain 2013]: http://dx.doi.org/10.12688/f1000research.2-191.v2
[@ouch]: http://cran.r-project.org/web/packages/ouch/
[@phylobase]: http://cran.r-project.org/web/packages/phylobase
[W3C 2013]: http://www.w3.org/TR/xhtml-rdfa-primer/ "RDFa 1.1 Primer - Second Edition: Rich Structured Data Markup for Web Documents"
[Tenopir et al. 2011]: http://doi.org/10.1371/journal.pone.0021101 "Data Sharing by Scientists: Practices and Perceptions"

>>>>

<!-- Emphasize: 
1) storing character data and tree data together, convenient for the comparative methods community
2) Adding metadata, including automated addition of metadata
3)

-->

@@

Mention of data sharing challenges and opportunities.  See [Drew et al 2013]

@@

The current stable version of the `RNeXML` package can be installed from
the CRAN library using the standard <!-- Actually not yet -->

@@

This allows us to include multiple trees and character matrices to a single NeXML file.  
<!-- We can add relevant metadata to each tree, such as the citation information 
     indicating the paper in which it was first described.  -->

@@

`RNeXML` automatically adds basic metadata by default such as a timestamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.  To adjust these and other common metadata fields that should 
be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
function.  This function does not assume any familiarity with the concepts
of namespaces or other more advanced metadata tools discussed later.  The
`add_basic_meta` takes an existing `nexml` object (or will create a new one
if none is provided, like the other `add_` functions), and any of the following
fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Additional fields may be added to subsequent versions, see
`?add_basic_metadata` for the most up-to-date information.  Here we create
an new `nexml` object with the basic metadata for the `bird.orders` data
provided in the `ape` package:

```{r}
data(bird.orders)
nex <- add_trees(bird.orders)
nex <- add_basic_meta(nex,
  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",

  description =" This data set describes the phylogenetic relationships of the
     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
     and Ahlquist inferred this phylogeny from an extensive number of
     DNA/DNA hybridization experiments. The ``tapestry'' reported by
     these two authors (more than 1000 species out of the ca. 9000
     extant bird species) generated a lot of debates.

     The present tree is based on the relationships among orders. The
     branch lengths were calculated from the values of Delta T50H as
     found in Sibley and Ahlquist (1990, fig. 353).",

  citation = "Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
     classification of birds: a study in molecular evolution. New
     Haven: Yale University Press.",

  creator = "Sibley, C. G. and Ahlquist, J. E.")
```

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.

For instance, misspellings or inconsistent use of taxonomic names in phylogenetic
data are a common source of frustration in reusing phylogenetic trees. By 
associating the taxonomic labels against a specific, centralized database
we can catch typographical errors and indicate more precisely which classification
we are appealing to in the event of conflicting taxonomic definitions.  In this 
example, we indicate that we intend the bird orders listed in the tree to correspond
to the NCBI definitions

```{r}
nex <- taxize_nexml(nex, "NCBI")
```

This function uses the `taxize` R library ([Chamberlain 2013]) to check each taxon
label against the NCBI database.  If a unique match is found, a metadata annotation
is added to the taxon providing the NCBI identification number to the taxonomic unit.
If no match is found, the user is warned to check for possible typographic errors in
the taxonomic labels provided.

Much richer metadata annotation is possible. Later we illustrate how metadata annotation
can be used to extend the base NeXML format to represent new forms of data while maintaining
compatability with any NeXML parser. The `RNeXML` package can be easily extended to support
helper functions such as `taxize_nexml` to add additional metadata without imposing a large
burden on the user.

@@

Common metadata can be accessed with a few dedicated functions:

```{r}
get_citation(S100)
```

@@

Which returns text from the otu element labels, typically used to define
taxonomic names, rather than text from explicit meta elements.

We can also access metadata at a specific level (or use `level=all`
to extract all meta elements in a list).  Here we show only the first
few results:

```{r} 
otu_meta <- get_metadata(S100, level="otu")
otu_meta[1:2]
```

This approach does not preserve the relationship between the metadata element and the 
`otu` it was annotating, and as such is most useful for quick visual exploration. 
Metadata values for specific elements or sets of elemetns can always be accessed using
the S4 subsetting mechanisms (see section below).

## RDF manipulation and SPARQL queries

Consider a better example than this one:

```{r}
rdf <- get_rdf(S100)
```

Query the rdf using XPath expressions:

```{r}
library(XML)
xpathSApply(rdf, "//dc:title", xmlValue) 
```

XPath is an expressive syntax for querying XML-structured data such
as RDF.  Queries using the XPath syntax are possible in R using the
XML library [Temple Lang 2013, XML].  (Note that XPath is far more
expressive for querying the metadata once the RDFa has been extracted,
since attribute values are now interpreted as namespaced XML nodes and
not just arbitrary text.)

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see CITE SOMETHING.  Meanwhile, we just introduce the sytnax
necessary to apply such queries to NeXML files from R.  First we load
the necessary R library and import the RDF-extracted metadata:

```{r}
library(rrdf)
saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
lib <- load.rdf("rdf_meta.xml")
```

We can then make arbitrary SPARQL queries directly from R.  For instance,
here is a simple query to extract the value of any `dc:title` element, analogous
to the XPath expression above:

```{r}
sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
```

@@

### Publishing NeXML files from R

Data archiving is increasingly required by journals in evolutionary biology and 
biodiversity [e.g. @Whitlock2011].  The burden of preparing and submitting properly
annotated data to archives continues to be a significant barrier [Tenopir et al. 2011, Stodden 2013],
and many phylogenetic trees upon which studies are based are inaccessible or lost 
to the research community [Drew et al 2013].  `RNeXML` seeks to lower these barriers
by providing functions to immediately archive NeXML files in scientific repositories.

```{r eval = FALSE}
id <- nexml_publish(nexml, visibility = "public", repository="figshare")
```

_Show downloading the nexml file directly from figshare, explain value
of global paths instead of personal directory paths appearing in scripts, etc_

Robust data archiving need not involve immediate release of the data. The `figshare` 
repository supports the option of `visibility = "private"`, allowing secure 
publication that provides a way to backup and share data with collaborators 
prior to public release.

<!-- Wouldn't it be nice to publish to Dryad or TreeBASE in this way?  -->

@@

[Temple Lang 2013, XML]: http://cran.r-project.org/web/packages/XML/
[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html
[Chamberlain 2013]: http://dx.doi.org/10.12688/f1000research.2-191.v2
[@ouch]: http://cran.r-project.org/web/packages/ouch/
[@phylobase]: http://cran.r-project.org/web/packages/phylobase
[W3C 2013]: http://www.w3.org/TR/xhtml-rdfa-primer/ "RDFa 1.1 Primer - Second Edition: Rich Structured Data Markup for Web Documents"
[Tenopir et al. 2011]: http://doi.org/10.1371/journal.pone.0021101 "Data Sharing by Scientists: Practices and Perceptions"
[Drew et al 2013]: http://doi.org/10.1371/journal.pbio.1001636
